Appendix 3: Estimating species interactions

### 2015-11-22 Harris_2015/manuscript-materials/draft.Rmd
almost done

\noindent
To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these
interactions in species composition data [@macarthur_population_1958;
@diamond_island_1975]. Despite decades of work and several major controversies,
however [@lewin_santa_1983; @strong_ecological_1984; @connor_checkered_2013],
existing methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
another can become lost in a web of indirect effects. For example, the
competitive interaction between the two shrub species in Figure 1A is
obscured by their shared tendency to occur in unshaded areas (Figure 1B). While
ecologists have long known that indirect effects can overwhelm direct ones at
the landscape level [@dodson_complementary_1970; @levine_competitive_1976], the
vast majority of our methods for drawing inferences from observational data do
not control for these effects [e.g. @diamond_island_1975;
@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
@pollock_understanding_2014]. To the extent that indirect interactions like
those in Figure 1 are generally important [@dodson_complementary_1970], existing
methods will not provide much evidence regarding species interactions.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it nevertheless does leave a
signal in the data (Figure 1C). For example, after partitioning the data set
into shaded and unshaded sites, there will be co-occurrence deficits in
each subset that wouldn't otherwise be apparent. More generally, controlling for
other species in the network will often be important for obtaining reliable
estimates of direct (conditional, or all-else-equal) effects.  This kind of
precision is difficult to obtain from null models, which only simulate the most
extreme possible scenario: that *all* direct and indirect interactions are
exactly zero. Nevertheless, null models have dominated this field for more than
three decades [@strong_ecological_1984; @gotelli_empirical_2009].

@@

\noindent
\textbf{Markov networks.}
Markov networks provide a framework for translating back and forth between the
conditional (all-else-equal) relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I show how
a set of conditional relationships can be used to determine how groups of species
can co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it controls the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends---all else equal---to
produce assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an ecologist
might expect). Of course, if all else is *not* equal (e.g. Figure 1, where the
presence of one competitor is associated with release from another competitor),
then predicting species' overall co-occurrence rates can be more complicated,
and may require summing over the different possible assemblages (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average [i.e., maximizing the likelihood ensures that the sufficient
statistics of the model distribution match the sufficient statistics of the data;
@murphy_machine_2012]. I used the rosalia package [@harris_rosalia_2015] for the
R programming language [@r_core_team_r_2015] to optimize the Markov network
parameters.

@@

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative interaction
metric was the sample correlation between species' presence-absence vectors,
which summarizes their marginal association.  Next, I used partial correlations,
which summarize species' conditional relationships. This approach is
common in molecular biology [@friedman_sparse_2008], but is rare in ecology
(see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions).
In the context of non-Gaussian data, the partial correlation can be thought of
as a computationally efficient approximation to the full Markov network model
[@loh_structure_2013]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network
[@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized
logistic regression models [@gelman_weakly_2008] for each species, using the
other species on the landscape as predictors. This produced two interaction
estimates for each species pair; one for the effect of species $i$ on species
$j$ and one for the reverse. These two coefficients are not identifiable from
the data, however [@schmidt_modeling_2012], so I used their average as an
overall measure of the overall relationship between species $i$ and species $j$.

@@

I then evaluated the relationship between each method's estimates and the "true"
interaction strengths among all of the species pairs from the larger simulated
landscapes. This determined which of the methods provide a consistent way to
know how strong species interactions are---regardless of which species were
present in a particular data set or how many observations were taken. To ensure
that the different test statistics (e.g. correlations versus $Z$ scores) were on
a common scale, I rescaled them using linear regression through the origin. I
then calculated the proportion of variance explained for different combinations
of model type and landscape size (compared with a null model that assumed all
interaction strengths to be zero).

@@

\noindent
\textbf{Twenty species.}
In general, each model's performance was highest for large landscapes with
simple assembly rules and no environmental heterogeneity (Figure 3).
Despite some variability across contexts, the rank ordering across methods was
very consistent.  In particular, the four methods that controlled for indirect
effects (the Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did not.  
Of these, the Markov network consistently performed the best. As anticipated by
@lee_learning_2012, generalized linear models closely
approximated the Markov network estimates, especially when the data sets were
very large [[Figure]]. As reviewed in @gotelli_empirical_2009, however, most analyses in
this field of ecology involve fewer than 50 sites, so model performance on the smaller
landscapes (where the gap was larger) is very important. As shown in Appendix
5, the standard errors associated with the estimates in Figure 3 are small
(less than 0.01), so the differences among methods should not be attributed to
sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's approximate Type I error rate (defined here
as the probability that 0 fell outside the 95% confidence interval for a pair of
species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.02 for
simulations that matched the model's assumptions, versus 0.14 for simulations
that included environmental heterogeneity (see Appendix 5 for confidence
interval coverage across a range of $\beta_{ij}$ values). The null model's
Type I error rates were [[XX]] and [[XX]] for the constant and heterogeneous
landscapse, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability that
the null model or the Markov network will predict the wrong sign of the
interaction with 95% confidence.  The null model makes such errors [[XX]] times
as often as the Markov network, even though it only reject the null hypothesis
[[XX]] as often overall (Appendix 5).  The Markov network's errors were also
more concentrated around 0, so it never misclassified strong interactions like
the null model did.

@@

Controlling for indirect effects via simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the range
of results shown in Figure 3, scientists in this field could more than double
their explanatory power by switching from null models to Markov networks (or
increase it nearly as much with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for indirect
effects could clearly improve our understanding of species' direct effects on
one another and on community structure.

@@

>>>>

\noindent
To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these
interactions in species composition data [@macarthur_population_1958;
@diamond_island_1975]. Despite decades of work and several major controversies,
however [@lewin_santa_1983; @strong_ecological_1984; @connor_checkered_2013],
existing methods for detecting competition's effects on community structure are
unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
another can become lost in a web of indirect effects. For example, the
competitive interaction between the two shrub species in Figure 1A is
obscured by their shared tendency to occur in unshaded areas (Figure 1B). While
ecologists have long known that indirect effects can overwhelm direct ones at
the landscape level [@levine_competitive_1976], the
vast majority of our methods for drawing inferences from observational data do
not control for these effects [e.g. @diamond_island_1975;
@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
@pollock_understanding_2014]. To the extent that indirect interactions like
those in Figure 1 are generally important, existing
methods will not provide much evidence regarding species interactions.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it does still leave a
signal in the data (Figure 1C). For example, after partitioning the data set
into shaded and unshaded sites, there will be co-occurrence deficits in
each subset that wouldn't otherwise be apparent. More generally, controlling for
other species in the network will often be important for obtaining reliable
estimates of direct (conditional, or all-else-equal) effects.  This kind of
precision is difficult to obtain from null models, which only test the most
extreme possible hypothesis: that *all* direct and indirect interactions are
exactly zero. Nevertheless, null models have dominated this field for more than
three decades [@strong_ecological_1984; @gotelli_empirical_2009].

@@

\noindent
\textbf{Markov networks.}
Markov networks provide a framework for translating back and forth between the
conditional (all-else-equal) relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I show how
a set of conditional relationships can determine species composition. Methods for
estimating conditional relationships from data are discussed in the next section.

@@

\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
species. In this model, $\alpha_{i}$ is an intercept term determining the amount
that the presence of species $i$ contributes to the log-probability of
$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
contributes to the log-probability; it controls the conditional relationship
between two species, i.e. the probability that they will be found together,
after controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
$e^2$ times higher when the other one is present (as compared with otherwise
equivalent sites). The relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur. As a result, the model tends---all else
equal---to produce assemblages where many positively-associated species pairs
co-occur and few negatively-associated pairs do (just as an ecologist might
expect). When all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then predicting
species' overall co-occurrence rates can be more complicated, and may require
summing over the different possible assemblages (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When the
number of species is reasonably small (less than about 30), one can find
exact maximum likelihood estimates for all of the $\alpha$ and $\beta$
coefficients given a presence-absence matrix by numerically
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will converge on the global maximum. I used the
rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its
gradients; the package passes these functions to the "BFGS" method in R's
general-purpose optimizer, which then estimates the Markov network parameters.

@@

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative interaction
metric was the sample correlation between species' presence-absence vectors,
which summarizes their marginal association.  Next, I used partial correlations,
which summarize species' conditional relationships. This approach is
common in molecular biology [@friedman_sparse_2008], but is rare in ecology
(see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions).
In the context of non-Gaussian data, the partial correlation can be thought of
as a computationally efficient approximation to the full Markov network model
[@loh_structure_2013]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate provided by the corpcor package’s `pcor.shrink` function with the
default settings [@schafer_corpcor_2014].

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network
[@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized
logistic regression models [@gelman_weakly_2008] for each species, using the
other species on the landscape as predictors. This produced two interaction
estimates for each species pair; one for the effect of species $i$ on species
$j$ and one for the reverse. These two coefficients are not identifiable from
the data, however [@schmidt_modeling_2012], so I used their average as an
overall measure of their relationship.

@@

For the larger landscapes, I evaluated the relationship between each method's
estimates and the "true" interaction strengths. To ensure that the different
test statistics (e.g. correlations versus $Z$ scores) were on a common scale, I
rescaled them using linear regression through the origin. I then calculated the
proportion of variance explained for different combinations of model type and
landscape size (compared with a null model that assumed all interaction
strengths to be zero).

@@

\noindent
\textbf{Twenty species.}
In general, each model's performance was highest for large landscapes with
simple assembly rules and no environmental heterogeneity (Figure 3).
Despite some variability across contexts, the rank ordering across methods was
very consistent.  In particular, the four methods that controlled for indirect
effects (the Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did not.
The Markov network consistently performed best of all. As anticipated by
@lee_learning_2012, generalized linear models closely
approximated the Markov network estimates, especially when the data sets were
very large [[Figure]]. As reviewed in @gotelli_empirical_2009, however, most analyses in
this field of ecology involve fewer than 50 sites, so model performance on the smaller
landscapes (where the gap was larger) is very important. As shown in Appendix
5, the standard errors associated with the estimates in Figure 3 are small
(less than 0.01), so the differences among methods should not be attributed to
sampling error.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, for which the true interaction rates are easiest to
interpret).  The Markov network's approximate Type I error rate (defined here
as the probability that 0 fell outside the 95% confidence interval for a pair of
species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.02 for
simulations that matched the model's assumptions, versus 0.14 for simulations
that included environmental heterogeneity (see Appendix 5 for confidence
interval coverage across a range of $\beta_{ij}$ values). The null model's
Type I error rates were [[XX]] and [[XX]] for the constant and heterogeneous
landscapes, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability that
the null model or the Markov network will predict the wrong sign of the
interaction with 95% confidence.  The null model makes such errors [[XX]] times
as often as the Markov network, even though it only reject the null hypothesis
[[XX]] as often overall (Appendix 5).  The Markov network's errors were also
more concentrated around 0, so it never misclassified strong interactions like
the null model did.

@@

Controlling for indirect effects via simultaneous estimation of multiple
ecological parameters seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the range
of results shown in Figure 3, scientists in this field could often triple
their explanatory power by switching from null models to Markov networks (or
increase it nearly as much with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for indirect
effects could clearly improve our understanding of species' direct effects on
one another and on community structure.

@@

<!--references go here-->
